Global search algorithms for one-dimensional multiextremal optimization problems with nonlinear constraints

### 2017-09-24 2017/ParOptBook/Chapter4.tex
parOptBook: update chapter 4

%% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\documentclass[graybox]{svmult}
%
%% choose options for [] as required from the list
%% in the Reference Guide
%
%\usepackage{mathptmx}       % selects Times Roman as basic font
%\usepackage{helvet}         % selects Helvetica as sans-serif font
%\usepackage{courier}        % selects Courier as typewriter font
%\usepackage{type1cm}        % activate if the above 3 fonts are
%                           % not available on your system
%
%\usepackage{makeidx}         % allows index generation
%\usepackage{graphicx}        % standard LaTeX graphics tool
%                             % when including figure files
%\usepackage{multicol}        % used for the two-column index
%\usepackage[bottom]{footmisc}% places footnotes at page bottom
%
%%custom package
%\usepackage{mathtools}
%
%% see the list of further useful packages
%% in the Reference Guide
%
%\makeindex             % used for the subject index
%                       % please use the style svind.ist with
%                       % your makeindex program
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\begin{document}

@@

\emph{Rule 5.} For all nonempty sets $I_\nu, \; 1 \leq \nu \leq m+1$, compute the estimates
\begin{equation}
  \label{eq4:mu_const}
  z_\nu^\ast = \left\{
  \begin{array}{lr}
    0, & \nu < M,\\
    \min\{ g_\nu(y(x_i): i\in I_\nu \}, & \nu = M.
  \end{array}
  \right.
\end{equation}

@@

\emph{Rule 8.} Make the next trial at the midpoint of the interval
$(x_{t-1},x_t)$ if the indices of the points $x_{t-1}$ and $x_t$  are not the same, i.e.
\[
x^{k+1} = \frac{x_t + x_{t-1}}{2}, \; \nu(x_{t-1}) \neq \nu(x_t).
\]
Otherwise, make the trial at the point
\begin{equation}
\label{eq4:next_point}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \frac{z_t-z_{t-1}}{2r_\nu\mu},\; \nu=\nu(x_{t-1})=\nu(x_t).
\end{equation}

The described rules can be supplemented with the termination condition, finalizing the iterations of the method if
\begin{equation}
\label{eq4:stop_cond}
  x_t - x_{t-1}\le\epsilon
\end{equation}
where t from \eqref{eq4:MaxR} and $\epsilon>0$ is the predefined accuracy of the optimization.

@@

The coordinates of the trials executed in the close points are marked by a dark rectangle. The values of the functions $g_1 , g_2 , g_3$ , and $\varphi=g_4$ have been computed $k_1=102$, $k_2=80, k_3=64,$ and $k_4=26$ times, respectively. As one can see in Fig. \ref{fig:4_7}, the trial points concentration takes place not only in the vicinity of the point of the global optimum $x^*=2,07957$, but also in the vicinities of the inadmissible boundary points of the domains $Q_j$ from \eqref{eq4:q}. This effect can be reduced by the substitution of estimates \eqref{eq4:mu_const} by the estimates
\begin{equation}
  \label{eq4:mu_const}
  z_\nu^\ast = \left\{
  \begin{array}{lr}
    -\epsilon_\nu, & \nu < M,\\
    \min\{ g_\nu(y(x_i): i\in I_\nu \}, & \nu = M.
  \end{array}
  \right.
\end{equation}
where $\epsilon_R=(\epsilon_1 ,\dots, \epsilon_m)$ is a predefined vector with positive components. Thus, $z^*=0$ if there exist the points $x_i,1\le i\le k$ from series \eqref{eq4:search_data} with the index greater than $\nu$. As a consequence of this substitution the values of the characteristics $R(i)$ from \eqref{eq4:characteristic} will become lower (because of the subtraction of $\epsilon_\nu$) if
\begin{equation}
  \nu=\max\{\nu(x_{i-1},\nu(x_i)\}<\max\{\nu(x_j):1\le j\le k\}\le V
\end{equation}
that makes executing the trials in the corresponding intervals  ($x_{i-1}$ , $x_i$ ) less probable. In all other aspects, the algorithm remains the same.

@@

The algorithm proposed below allows the modification of the constraint order check during solving the problem \eqref{eq4:problem}. These modifications are firstly directed onto analyzing during the current iteration the constraint violated at the end of the interval including the current iteration point. As it has been already mentioned above, such an approach might reduce the number of checks at the new trial point since the violation of this constraint in a vicinity of already found violation might be quite probable. Hence, the need in new rules for the juxtaposition of the indices to the points from \eqref{eq4:search_data} (remind, that these indices determine the use of expressions \eqref{eq4:characteristic} for the computation of the characteristics) as well as in a special order
of the constraint check at the current trial point arises. Note that the formation of such order is inessential for the intervals, the ends of which belong to a feasible domain $Q$ from (4.31) since a necessity for checking all constraints is probable in such intervals.

Assume the constraint with the number $j,\: 1\le j\le m$, to fulfill at all points of the domain
\begin{equation}
  Q_j=\{x\in[a,b]:g_j(x)\le 0\},\: 1\le j\le m,
\end{equation}
which hereinafter will be called the feasible domain for this constraint. The feasible domain $Q$ of initial problem \eqref{eq4:problem} can be defined by the condition
\begin{equation}
  Q=Q_1\cap\dots\cap Q_m.
\end{equation}

@@

\begin{theorem}
Assume problem \eqref{eq4:problem} to have a solution $x^*$ and the following conditions are fulfilled:
\begin{enumerate}
  \item each domain $Q_j ,1\le j\le m+1$ from (4.30) represents a union of a finite number of segments with a positive length;
  \item each function $g_j(x): 1\le j\le m+1$ satisfies Lipschitz condition with the corresponding constant $L_j$;
  \item the components $\epsilon_\nu$ of the reserve vector $\epsilon_R$ satisfy the inequalities
  \begin{equation}
    0<2\epsilon_\nu<L_\nu(\beta-\alpha),
  \end{equation}
  where $\beta-\alpha$ is the length of the interval $[\alpha,\beta]$ belonging to the feasible domain $Q$ from (4.31) and including the point $x^*$, which is the solution of the problem;
  \item since some sufficiently large step number $k$ the values $\mu_\nu$ from (4.15) corresponding to the nonempty sets $I_\nu$ satisfy the inequalities
  \begin{displaymath}
    r_\nu\mu_\nu>2L_\nu,\: 1\le\nu\le m+1,
  \end{displaymath}
  where $r_\nu>1,\: 1\le\nu\le m+1$ are the parameters of the algorithm.
\end{enumerate}

@@

\begin{figure}[h]
  \label{fig:4_17}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_16.png}
  \caption{Average number of function values computations for the class $G_{HL}$}
\end{figure}

@@

>>>>

% RECOMMENDED %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[graybox]{svmult}

% choose options for [] as required from the list
% in the Reference Guide

\usepackage{mathptmx}       % selects Times Roman as basic font
\usepackage{helvet}         % selects Helvetica as sans-serif font
\usepackage{courier}        % selects Courier as typewriter font
\usepackage{type1cm}        % activate if the above 3 fonts are
                           % not available on your system

\usepackage{makeidx}         % allows index generation
\usepackage{graphicx}        % standard LaTeX graphics tool
                             % when including figure files
\usepackage{multicol}        % used for the two-column index
\usepackage[bottom]{footmisc}% places footnotes at page bottom

%custom package
\usepackage{mathtools}

% see the list of further useful packages
% in the Reference Guide

\makeindex             % used for the subject index
                       % please use the style svind.ist with
                       % your makeindex program

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

@@

\emph{Rule 5.} For all nonempty sets $I_\nu, \; 1 \leq \nu \leq m+1$, compute the estimates
\begin{equation}
  \label{eq4:z_const}
  z_\nu^\ast = \left\{
  \begin{array}{lr}
    0, & \nu < M,\\
    \min\{ g_\nu(y(x_i): i\in I_\nu \}, & \nu = M.
  \end{array}
  \right.
\end{equation}

@@

\emph{Rule 8.} Make the next trial at the midpoint of the interval
$(x_{t-1},x_t)$ if the indices of the points $x_{t-1}$ and $x_t$  are not the same, i.e.
\[
x^{k+1} = \frac{x_t + x_{t-1}}{2}, \; \nu(x_{t-1}) \neq \nu(x_t).
\]
Otherwise, make the trial at the point
\begin{equation}
\label{eq4:next_point}
x^{k+1} = \frac{x_t+x_{t-1}}{2} - \frac{z_t-z_{t-1}}{2r_\nu\mu_\nu},\; \nu=\nu(x_{t-1})=\nu(x_t).
\end{equation}

The described rules can be supplemented with the termination condition, finalizing the iterations of the method if
\begin{equation}
\label{eq4:stop_cond}
  x_t - x_{t-1}\le\epsilon
\end{equation}
where $t$ from \eqref{eq4:MaxR} and $\epsilon>0$ is the predefined accuracy of the optimization.

@@

The coordinates of the trials executed in the close points are marked by a dark rectangle. The values of the functions $g_1 , g_2 , g_3$ , and $\varphi=g_4$ have been computed $k_1=102$, $k_2=80, k_3=64,$ and $k_4=26$ times, respectively. As one can see in Fig. \ref{fig:4_7}, the trial points concentration takes place not only in the vicinity of the point of the global optimum $x^*=2,07957$, but also in the vicinities of the inadmissible boundary points of the domains $Q_j$ from \eqref{eq4:q}. This effect can be reduced by the substitution of estimates \eqref{eq4:z_const} by the estimates
\begin{equation}
  z_\nu^\ast = \left\{
  \begin{array}{lr}
    -\epsilon_\nu, & \nu < M,\\
    \min\{ g_\nu(y(x_i): i\in I_\nu \}, & \nu = M.
  \end{array}
  \right.
\end{equation}
where $\epsilon_R=(\epsilon_1 ,\dots, \epsilon_m)$ is a predefined vector with positive components. Thus, $z^*=0$ if there exist the points $x_i,1\le i\le k$ from series \eqref{eq4:search_data} with the index greater than $\nu$. As a consequence of this substitution the values of the characteristics $R(i)$ from \eqref{eq4:characteristic} will become lower (because of the subtraction of $\epsilon_\nu$) if
\begin{equation}
  \nu=\max\{\nu(x_{i-1},\nu(x_i)\}<\max\{\nu(x_j):1\le j\le k\}\le V
\end{equation}
that makes executing the trials in the corresponding intervals  ($x_{i-1}$ , $x_i$ ) less probable. In all other aspects, the algorithm remains the same.

@@

The algorithm proposed below allows the modification of the constraint order check during solving the problem \eqref{eq4:problem}. These modifications are firstly directed onto analyzing during the current iteration the constraint violated at the end of the interval including the current iteration point. As it has been already mentioned above, such an approach might reduce the number of checks at the new trial point since the violation of this constraint in a vicinity of already found violation might be quite probable. Hence, the need in new rules for the juxtaposition of the indices to the points from \eqref{eq4:search_data} (remind, that these indices determine the use of expressions \eqref{eq4:characteristic} for the computation of the characteristics) as well as in a special order
of the constraint check at the current trial point arises. Note that the formation of such order is inessential for the intervals, the ends of which belong to a feasible domain $Q$ from \eqref{eq4:Q_union} since a necessity for checking all constraints is probable in such intervals.

Assume the constraint with the number $j,\: 1\le j\le m$, to fulfill at all points of the domain
\begin{equation}
  Q_j=\{x\in[a,b]:g_j(x)\le 0\},\: 1\le j\le m,
\end{equation}
which hereinafter will be called the feasible domain for this constraint. The feasible domain $Q$ of initial problem \eqref{eq4:problem} can be defined by the condition
\begin{equation}
  \label{eq4:Q_union}
  Q=Q_1\cap\dots\cap Q_m.
\end{equation}

@@

\begin{theorem}
Assume problem \eqref{eq4:problem} to have a solution $x^*$ and the following conditions are fulfilled:
\begin{enumerate}
  \item each domain $Q_j ,1\le j\le m+1$ from (4.30) represents a union of a finite number of segments with a positive length;
  \item each function $g_j(x): 1\le j\le m+1$ satisfies Lipschitz condition with the corresponding constant $L_j$;
  \item the components $\epsilon_\nu$ of the reserve vector $\epsilon_R$ satisfy the inequalities
  \begin{equation}
    0<2\epsilon_\nu<L_\nu(\beta-\alpha),
  \end{equation}
  where $\beta-\alpha$ is the length of the interval $[\alpha,\beta]$ belonging to the feasible domain $Q$ from \eqref{eq4:Q_union} and including the point $x^*$, which is the solution of the problem;
  \item since some sufficiently large step number $k$ the values $\mu_\nu$ from (4.15) corresponding to the nonempty sets $I_\nu$ satisfy the inequalities
  \begin{displaymath}
    r_\nu\mu_\nu>2L_\nu,\: 1\le\nu\le m+1,
  \end{displaymath}
  where $r_\nu>1,\: 1\le\nu\le m+1$ are the parameters of the algorithm.
\end{enumerate}

@@

\begin{figure}[h]
  \label{fig:4_17}
  \centering
  \includegraphics[width=0.8\textwidth]{figures/4_17.png}
  \caption{Average number of function values computations for the class $G_{HL}$}
\end{figure}

@@

The first $p$ trial points are selected at the internal points of the interval $[a,b]$ arbitrarily. Each processor starts the trials execution at the corresponding points. After completing the $k^{th}$ trial, when the points $x^{k+1},\dots, x^{k+p - 1}$ have already been selected but the trial results at these points are not obtained yet, the point of the next trial $x^{k+p}$ is selected according to the following rules.

\emph{Rule 1.} Renumber the points of the preceding trials $x^1,\dots,x^k$ by the subscripts in the increasing order of the coordinate values, i.e.,
\begin{equation}
  \label{eq4:par_search_data}
  a=x_0<x_1<\dots x_i<\dots<x_{k+p-1}<x_{k+p}=b,
\end{equation}
the points $x_0=a$ and $x_{k + p}=b$ are included into series \eqref{eq4:par_search_data} additionally for the convenience of the future notation.

\emph{Rule 2.} Determine the set $I$ of the points $x_i,\: 1\le i<k+p$, such that the trials at these points have not been completed yet.

\emph{Rule 3.} Juxtapose to each point $x_i,\: 1\le i<k+p$, from series \eqref{eq4:par_search_data} such that $x_i\not\in I$ the
index $\nu=\nu(x_i)$ and the value $z_i=g_\nu(x_i )$ calculated at this point (the values $z_0$ and $z_{k + p}$ are not determined).

\emph{Rule 4.} Construct the sets
\begin{equation}
  I_\nu=\{i:1\le i<k+p, \nu=\nu(x_i),x_i\not\in I\},\:1\le\nu\le m+1,
\end{equation}
including the subscripts of all points $x_i ,1\le i<k+p$, with the same index, in which the trials are completed at the moment. The boundary points $x_0=a$ and $x_{k + 1}=b$ are considered as the points with the zero indices. The additional set $I_0=\{0,k+1\}$ is associated with these points.

Find the maximum value of the index
\begin{equation}
  M=\max\{\nu=\nu(x_i),x_i\not\in I,1\le i \le k\}.
\end{equation}

\emph{Rule 5.} Compute current estimates
\begin{equation}
  \label{eq4:lip_est_par}
  \mu = \max\left\{ \frac{\left|z_i-z_j\right|}{ x_i - x_j }, \; i,j \in I_\nu, \; i>j \right\},
\end{equation}
for the unknown Lipschitz constants $L_\nu$ of the functions $g_\nu,\: 1\le\nu\le m+1$. If the set $I_\nu$ includes less than two elements or if $\mu_\nu$ from \eqref{eq4:lip_est_par} appears to be equal to zero, to set $\mu_\nu=1$.

\emph{Rule 6.} For all nonempty sets $I_\nu,\: 1\le\nu\le m+1$ calculate the estimates
\begin{equation}
  \label{eq4:z_const_par}
  z_\nu^\ast = \left\{
  \begin{array}{lr}
    -\epsilon_\nu, & \nu < M,\\
    \min\{ g_\nu(y(x_i): i\in I_\nu \}, & \nu = M.
  \end{array}
  \right.
\end{equation}
where the vector $\epsilon_R=(\epsilon_1 ,\dots, \epsilon_m )$ is a given vector with the non-negative components (\emph{the reserve vector}).

\emph{Rule 7.} Compute the characteristics $R(i)$ for each interval $(x_{i-1}, x_i ),\: x_{i – 1}\not\in I,\: x_i\not\in I, 1\le i\le k+p$ where
\begin{gather}
  R(i)=2\Delta_i-4\frac{z_i-z_\nu^\ast}{r_\nu \mu_\nu}, \; \nu=\nu(x_i)>\nu(x_{i-1}), \nonumber \\
  \label{eq4:characteristic_par}
  R(i)=\Delta_i+\frac{(z_i-z_{i-1})^2}{r_\nu^2 \mu_\nu^2\Delta_i}-2\frac{z_i+z_{i-1}-2z_\nu^\ast}{r_\nu \mu_\nu}, \;  \nu=\nu(x_i)=\nu(x_{i-1}),\\
  R(i)=2\Delta_i-4\frac{z_{i-1}-z_\nu^\ast}{r_\nu \mu_\nu}, \; \nu=\nu(x_{i-1})>\nu(x_i), \nonumber \\
  \Delta_i=x_i - x_{i-1} \nonumber
\end{gather}
where the values $r_\nu>1,\: 1\le\nu\le m+1$ are the algorithm parameters and $\Delta_i=x_i-x_{i-1}$.

\emph{Rule 8.} Select the interval $(x_{t-1},x_t )$ whith the largest characteristic
\begin{equation}
  \label{eq4:MaxR_par}
  R(t)=\max\{R(i):x_{i-1}\not\in I,x_{i}\not\in I,1\le i\le k+p\}.
\end{equation}

\emph{Rule 9.} Execute the next trial at the middle point of the interval $(x_{t-1},x_t )$ if the indices of its boundary points are not the same, i.e.,
\begin{equation}
  x^{k+1} = \frac{x_t + x_{t-1}}{2}, \; \nu(x_{t-1}) \neq \nu(x_t).
\end{equation}
Otherwise, accept
\begin{equation}
\label{eq4:next_point_par}
x^{k+p} = \frac{x_t+x_{t-1}}{2} - \frac{z_t-z_{t-1}}{2r_\nu\mu_\nu},\; \nu=\nu(x_{t-1})=\nu(x_t).
\end{equation}
as the point of the next trial.

The rules described above can be supplemented with the termination condition
\begin{equation}
\label{eq4:stop_cond_par}
  x_t - x_{t-1}\le\epsilon
\end{equation}
where $t$ from \eqref{eq4:MaxR_par} and $\epsilon>0$ is the predefined accuracy of the search.

\subsection{The convergence conditions of the algorithm}
\begin{theorem}
  \label{th4:convergence_par}
  Assume problem \eqref{eq4:problem} to have a solution $x^*$ and the following conditions to be fulfilled:
  \begin{itemize}
    \item each subdomain $Q_j ,1\le j\le m+1$ from \eqref{eq4:q} is a union of a finite number of positive length segments;
    \item each function $g_j ,1\le j\le m+1$ satisfies Lipschitz condition with the corresponding constant $L_j$;
    \item the components $\epsilon_\nu$ of the reserve vector $\epsilon_R$ satisfy the inequalities
    \begin{equation}
      \label{eq4:reserves_par}
      0<2\epsilon_\nu<L_\nu(\beta-\alpha)
    \end{equation}
    where $\beta-\alpha$ is the length of the interval $[\alpha,\beta]$ belonging to the feasible domain $Q$ from \eqref{eq4:Q_union} and including the point $x^*$ which is the solution of the problem;
    \item beginning from sufficiently large step number $k$, the values $\nu_\nu$ from \eqref{eq4:lip_est_par} corresponding to the nonempty sets $I_\nu$ satisfy the inequalities
    \begin{equation}
      r_\nu\mu_\nu>2L_\nu,\:1\le\nu\le m+1,
    \end{equation}
    where the algorithm parameters $r_\nu>1,\:1\le\nu\le m+1$;
    \item during the parallelization, the inequalities
    \begin{displaymath}
      T^j_{l+1}-T^j_j<T<\infty,1\le j\le p, l>1,
    \end{displaymath}
    are true where $T_l^j$ is the moment of completing the $l$ th trial by the $j^{th}$ processor and the inequality $1<p<P<\infty$ takes place for the number of processors $p$.
  \end{itemize}
  Then:
  \begin{itemize}
    \item the point $x^*$ is a limit one for the sequence ${x^k }$ generated by the described asynchronous algorithm for problem \eqref{eq4:problem} if $\epsilon=0$ in termination condition \eqref{eq4:stop_cond_par};
    \item any limit point $\overline x$ of the sequence $\{x_k \}$ is a solution of the problem \eqref{eq4:problem};
    \item the convergence to the limit point $\overline x$ is bilateral if $\overline x\not=a$ and $\overline x\not=b$.
  \end{itemize}
\end{theorem}

Before the proof of the theorem, let us formulate and prove several auxiliary statements.

\begin{lemma}
  (on the sequence of the shrinking intervals). Assume the conditions of Theorem \ref{th4:convergence_par} to be satisfied. Then for each limit point $x$ of the sequence $\{x_k\}$ there exists an infinite sequence of the intervals
  \begin{equation}
    \{(x_{t-1},x_t):t=t(q^s)\},\:s=1,2,\dots,
  \end{equation}
  satisfying the conditions
  \begin{gather}
    \overline x= \cap_{s=1}^\infty[x_{t-1},x_t], \\
    \lim_{p\to\infty}\Delta_t=0, \\
    R(t(q^s))>0,\:s=1,2,\dots, \\
    \lim_{p\to\infty}R(t(q^s))=0,
  \end{gather}
  where $q^1<q^2<\dots;\:R(t),\:\Delta_t$, and $t$ are from rules \eqref{eq4:characteristic_par} and \eqref{eq4:MaxR_par}, correspondingly.
\end{lemma}

\begin{proof}
  Since $\overline x$ is a limit point, there exists a sequence of the trial numbers $q^1,q^2,\dots$ satisfying the conditions
  \begin{equation}
    x^{q+1},\overline x\in[x_{t-1},x_t], t=t(q),q\in\{q^p\},p=1,2,\dots.
  \end{equation}
  These conditions reflect the fact that the point of the $(q+1)^{th}$ trial falls into the interval $[x_{t-1},x_t]$ containing the point $\overline x$ at the step $q$, i.e., $t=t(q)$.
\end{proof}
